Exploring mathematical objects from custom-tailored mathematical universes

### 2020-01-14 paper-filmat.tex
Details on the Zariski topos

\documentclass[oneside]{amsart}

@@

\title[]{Exploring mathematical objects from custom-tailored mathematical universes}
\author{Ingo Blechschmidt}
\address{Università di Verona \\
Department of Computer Science \\
Strada le Grazie 15 \\
37134 Verona, Italy}
\email{iblech@speicherleck.de}

@@

\newcommand{\xra}[1]{\xrightarrow{#1}}
\newcommand{\XXX}[1]{\textbf{XXX: #1}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\ZZ}{\mathbb{Z}}
\renewcommand{\P}{\mathcal{P}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\op}{\mathrm{op}}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\Mod}{Mod}
\DeclareMathOperator{\Sh}{Sh}
\DeclareMathOperator{\PSh}{PSh}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\renewcommand{\_}{\mathpunct{.}\,}
\newcommand{\effective}{ef{}fective\xspace}
\newcommand{\?}{\,{:}\,}
\newcommand{\realizes}{\Vdash}
\newcommand{\notnot}{\emph{not~not}\xspace}

@@

Secondly, there is a shift in emphasis. An important philosophical objective
for studying models of set theory is to explore which notions of sets are
coherent: Does the cardinality of the reals need to be the cardinal directly
succeeding~$\aleph_0$, the cardinality of the naturals? No, there are models of
set theory in which the continuum hypothesis fails. Do non-measurable sets of
reals need to exist? No, in models of~$\textsc{zf}+\textsc{ad}$,
Zermelo--Fraenkel set theory plus the axiom of determinacy, it is a theorem that
every subset of~$\RR^n$ is Lebesgue-measurable. Can the axiom of choice be
added to the axioms of~\textsc{zf} without causing inconsistency? Yes, if~$M$
is a model of~\textsc{zf} then~$L^M$, the set of definable sets of~$M$, is a
model of~\textsc{zfc}.  %~\cite{XXX, maybe SEP?}

@@

\begin{table}
  \begin{tabbing}
    $U \models (\forall x\?\RR\_ \varphi(x))$ \= \kill
    $\Sh(X) \models \varphi$ \> iff $X \models \varphi$. \\\\
    $U \models a = b$ \> iff~$a = b$ on~$U$. \\
    $U \models \top$ \> is true for any open~$U$. \\
    $U \models \bot$ \> iff~$U$ is the empty open. \\
    $U \models (\varphi \wedge \psi)$ \> iff~$U \models \varphi$ and~$U \models \psi$. \\
    $U \models (\varphi \vee \psi)$ \> iff there is an open covering~$U =
    \bigcup_i U_i$ such that, \\ \> \qquad for each~$i$, $U_i \models \varphi$
    or $U_i \models \psi$. \\
    $U \models (\varphi \Rightarrow \psi)$ \> iff for every open~$V \subseteq U$,
    $V \models \varphi$ implies~$V \models \psi$. \\
    $U \models (\forall a\?\RR\_ \varphi(x))$ \> iff for every open~$V
    \subseteq U$ and any continuous function~$a_0 : V \to \RR$, $V \models
    \varphi(a_0)$. \\
    $U \models (\exists a\?\RR\_ \varphi(x))$ \> iff there is an open
    covering~$U = \bigcup_i U_i$ such that, \\ \> \qquad for each~$i$,
    there exists a continuous function~$a_0 : U_i \to \RR$ with~$U_i \models
    \varphi(a_0)$.
  \end{tabbing}

@@

For instance, employing numbers~$\varepsilon$ such that~$\varepsilon^2 = 0$, we can
compute derivatives blithely as follows, without requiring the notion of
limits:
\begin{align*}
  (x + \varepsilon)^2 - x^2 &= x^2 + 2x\varepsilon + \varepsilon^2 - x^2 = 2x\varepsilon \\
  (x + \varepsilon)^3 - x^3 &= x^3 + 3x^2\varepsilon + 3x\varepsilon^2 + \varepsilon^3 - x^3 = 3x^2\varepsilon
\end{align*}
In each case, the derivative is visible as the coefficient of~$\varepsilon$ in
the result. A further example is from geometry: Having a nontrivial
set~$\Delta$ of infinitesimal numbers available allows us to define a
\emph{tangent vector} to a manifold~$M$ to be a map~$\gamma : \Delta \to M$. This
definition precisely captures the intuition that a tangent vector is an
infinitesimal curve.

@@

The following section presents the basics of \emph{synthetic differential
geometry}~\cite{kock:sdg,kock:new-methods}. This is a further topos-theoretic
approach to infinitesimals which is suited to illustrate the philosophy of toposes
as lenses. Its stated goal is to devise a rigorous context in which the
writings of Sophus Lie, who freely employed infinitesimals in his seminal work
on differential geometry, can be effortlessly interpreted, staying close to the
original.

\subsection{XXX ...}

@@

* use the phrase "very rich and very deep" (or similarly) about set theory
* PA doesn't have models
* Sven(?)

>>>>

\documentclass[oneside,reqno]{amsart}

@@

\title[]{Exploring mathematical objects from custom-tailored mathematical universes}
\author{Ingo Blechschmidt}
\address{Universität Augsburg \\
Institut für Mathematik \\
Universitätsstr. 14 \\
86159 Augsburg, Germany}
\email{ingo.blechschmidt@math.uni-augsburg.de}

@@

\newenvironment{indentblock}{%
  \list{}{\leftmargin\leftmargin}%
  \item\relax
}{%
  \endlist
}

\newcommand{\xra}[1]{\xrightarrow{#1}}
\newcommand{\XXX}[1]{\textbf{XXX: #1}}
\newcommand{\aaa}{\mathfrak{a}}
\newcommand{\bbb}{\mathfrak{b}}
\newcommand{\mmm}{\mathfrak{m}}
\newcommand{\I}{\mathcal{I}}
\newcommand{\J}{\mathcal{J}}
\newcommand{\E}{\mathcal{E}}
\newcommand{\F}{\mathcal{F}}
\newcommand{\B}{\mathcal{B}}
\newcommand{\NN}{\mathbb{N}}
\newcommand{\RR}{\mathbb{R}}
\newcommand{\ZZ}{\mathbb{Z}}
\renewcommand{\P}{\mathcal{P}}
\renewcommand{\O}{\mathcal{O}}
\newcommand{\defeq}{\vcentcolon=}
\newcommand{\op}{\mathrm{op}}
\DeclareMathOperator{\Spec}{Spec}
\DeclareMathOperator{\Hom}{Hom}
\DeclareMathOperator{\Mod}{Mod}
\DeclareMathOperator{\Sh}{Sh}
\DeclareMathOperator{\Zar}{Zar}
\DeclareMathOperator{\PSh}{PSh}
\newcommand{\Set}{\mathrm{Set}}
\newcommand{\Eff}{\mathrm{Ef{}f}}
\renewcommand{\_}{\mathpunct{.}\,}
\newcommand{\effective}{ef{}fective\xspace}
\newcommand{\?}{\,{:}\,}
\newcommand{\realizes}{\Vdash}
\newcommand{\notnot}{\emph{not~not}\xspace}

@@

Secondly, there is a shift in emphasis. An important philosophical objective
for studying models of set theory is to explore which notions of sets are
coherent: Does the cardinality of the reals need to be the cardinal directly
succeeding~$\aleph_0$, the cardinality of the naturals? No, there are models of
set theory in which the continuum hypothesis fails. Do non-measurable sets of
reals need to exist? No, in models of~$\textsc{zf}+\textsc{ad}$,
Zermelo--Fraenkel set theory plus the axiom of determinacy, it is a theorem that
every subset of~$\RR^n$ is Lebesgue-measurable. Can the axiom of choice be
added to the axioms of~\textsc{zf} without causing inconsistency? Yes, if~$M$
is a model of~\textsc{zf} then~$L^M$, the constructible sets of~$M$ form a
model of~\textsc{zfc}.  %~\cite{XXX, maybe SEP?}

@@

\begin{table}
  \begin{tabbing}
    $U \models (\forall x\?\RR\_ \varphi(x))$ \= \kill
    $\Sh(X) \models \varphi$ \> iff $X \models \varphi$. \\\\
    $U \models a = b$ \> iff~$a = b$ on~$U$. \\
    $U \models \top$ \> is true for any open~$U$. \\
    $U \models \bot$ \> iff~$U$ is the empty open. \\
    $U \models (\varphi \wedge \psi)$ \> iff~$U \models \varphi$ and~$U \models \psi$. \\
    $U \models (\varphi \vee \psi)$ \> iff there is an open covering~$U =
    \bigcup_i U_i$ such that, \\ \> \qquad for each index~$i$, $U_i \models \varphi$
    or $U_i \models \psi$. \\
    $U \models (\varphi \Rightarrow \psi)$ \> iff for every open~$V \subseteq U$,
    $V \models \varphi$ implies~$V \models \psi$. \\
    $U \models (\forall a\?\RR\_ \varphi(a))$ \> iff for every open~$V
    \subseteq U$ and \\\>\qquad any continuous function~$a_0 : V \to \RR$, $V \models
    \varphi(a_0)$. \\
    $U \models (\exists a\?\RR\_ \varphi(a))$ \> iff there is an open
    covering~$U = \bigcup_i U_i$ such that, \\ \> \qquad for each index~$i$,
    there exists a \\\>\qquad continuous function~$a_0 : U_i \to \RR$ with~$U_i \models
    \varphi(a_0)$.
  \end{tabbing}

@@

For instance, employing numbers~$\varepsilon$ such that~$\varepsilon^2 = 0$, we can
compute derivatives blithely as follows, without requiring the notion of
limits:
\begin{equation}\label{eq:derivative}\tag{$\star$}
  \begin{aligned}
    (x + \varepsilon)^2 - x^2 &= x^2 + 2x\varepsilon + \varepsilon^2 - x^2 = 2x\varepsilon \\
    (x + \varepsilon)^3 - x^3 &= x^3 + 3x^2\varepsilon + 3x\varepsilon^2 + \varepsilon^3 - x^3 = 3x^2\varepsilon
  \end{aligned}
\end{equation}
In each case, the derivative is visible as the coefficient of~$\varepsilon$ in
the result. A further example is from geometry: Having a nontrivial
set~$\Delta$ of infinitesimal numbers available allows us to define a
\emph{tangent vector} to a manifold~$M$ to be a map~$\gamma : \Delta \to M$. This
definition precisely captures the intuition that a tangent vector is an
infinitesimal curve.

@@

The following section presents a topos used in \emph{synthetic differential
geometry}~\cite{kock:sdg,kock:new-methods}. This is a further topos-theoretic
approach to infinitesimals which is suited to illustrate the philosophy of toposes
as lenses. Its stated goal is to devise a rigorous context in which the
writings of Sophus Lie, who freely employed infinitesimals in his seminal
works, can be effortlessly interpreted, staying close to the
original and requiring no coding.

\subsection{The Zariski topos} The starting point is the observation that
while the field~$\RR$ of ordinary real numbers doesn't contain infinitesimals
(except for zero), the ring~$\RR[\varepsilon]/(\varepsilon^2)$ of \emph{dual
numbers} does. This ring has the cartesian product~$\RR \times \RR$ as its
underlying set and the ring operations are defined such
that~$\varepsilon^2 = 0$, where~$\varepsilon \defeq \langle 0,1 \rangle$:
\[
  \langle a,b \rangle + \langle a',b' \rangle \defeq
  \langle a+a', b+b' \rangle \qquad
  \langle a,b \rangle \cdot \langle a',b' \rangle \defeq
  \langle aa', ab'+a'b \rangle
\]
We write~$\langle a,b \rangle$ more clearly
as~$a+b\varepsilon$.

The dual numbers contain infinitesimal numbers, more precisely \emph{nilsquare
numbers}, namely all numbers of the form~$b \varepsilon$ for~$b \in \RR$, and
they are sufficient to rigorously reproduce derivative computations of
polynomials such as~\eqref{eq:derivative}.

However, the dual numbers are severely lacking in other aspects. Firstly, they
don't contain any nilcubic numbers which are not already nilsquare. These are
required in order to extend calculations like~\eqref{eq:derivative} to second
derivatives, as in
\[ (x+\varepsilon)^3 - x^3 = x^3 + 3x^2\varepsilon + \frac{1}{2!} 6x
\varepsilon^2. \]
Secondly, the dual numbers contain, up to scaling, only a single infinitesimal
number. Further independent infinitesimals are required in order to deal with
functions of several variables, as in
\[ f(x+\varepsilon,y+\varepsilon') - f(x,y) = D_xf(x,y)\varepsilon +
D_yf(x,y)\varepsilon'. \]
Thirdly, and perhaps most importantly, the ring of dual numbers fails to be a
field. The only invertible dual numbers are the numbers of the form~$a +
b\varepsilon$ with~$a$ invertible in the reals; it is not true that any nonzero
dual number is invertible.

The first deficiency could be fixed by passing
from~$\RR[\varepsilon]/(\varepsilon^2)$ to~$\RR[\varepsilon]/(\varepsilon^3)$
(a ring whose elements are triples and whose ring operations are defined such
that~$\langle0,0,1\rangle^3 = 0$) and the second by passing
from~$\RR[\varepsilon]/(\varepsilon^2)$
to~$\RR[\varepsilon,\varepsilon']/(\varepsilon^2,\varepsilon'^2,\varepsilon\varepsilon')$.
In a sense, both of these proposed replacements are better \emph{stages} than
the basic~$\RR[\varepsilon]/(\varepsilon)^2$ or even~$\RR$ itself. However,
similar criticisms can be mounted against any of these better stages, and the
problem that all these substitutes are not fields persists.

\bigskip
\paragraph{Introducing the topos}
The \emph{Zariski topos of~$\RR$},~$\Zar(\RR)$, meets all of these challenges. It
contains a ring~$R$, the so-called \emph{ring of smooth numbers}, which reifies
the real numbers, the dual numbers, the two proposed better stages and indeed
any finitely presented~$\RR$-algebra into a single coherent entity. The
Kripke--Joyal translations rules of~$\Zar(\RR)$ are listed in
Table~\ref{table:zariski}. Any evaluation of an internal statement starts out
with the most basic stage of all, the ordinary reals~$\RR$; then, during the
course of evaluation, the current stage is successively refined to better
stages (further finitely presented~$\RR$-algebras).

For instance, universal quantification~``$\forall x \? R$'' not only refers to
all elements of the current stage, but also to any elements of any refinement
of the current stage. Similarly, negation~``$\neg\varphi$'' does not only mean
that~$\varphi$ would imply~$1 = 0$ in the current stage, but also that it does
so at any later stage.

For reference purposes only, we include the precise definition of the Zariski topos.
\begin{defn}The \emph{Zariski topos of~$\RR$},~$\Zar(\RR)$, is the category of
functors from finitely presented~$\RR$-algebras to sets. The object~$R$
of~$\Zar(\RR)$ is the tautologous functor~$A \mapsto A$.
\end{defn}

\paragraph{Properties of the smooth numbers}
As a concrete example, the Kripke--Joyal translation of the statement that~$R$ is a field,
\[ \Zar(\RR) \models \forall x\?R\_ (\neg(x = 0) \Rightarrow \exists y\?R\_ xy = 1), \]
is this:
\begin{indentblock}
For any stage~$A$ and any element~$x \in A$, \\
${\qquad}$ if for any later stage~$B$ of~$A$ \\
${\qquad\qquad}$ in which~$x = 0$ holds \\
${\qquad\qquad}$ also~$1 = 0$ holds, \\
${\qquad}$ then~$A$ can be covered by later stages~$B_i$ such that, for each
index~$i$, \\
${\qquad\qquad}$ there is an element~$y \in B_i$ such that~$xy = 1$ in~$B_i$.
\end{indentblock}
And indeed, this statement is true. Let a stage~$A$ (a finitely
presented~$\RR$-algebra) and an element~$x \in A$ which fulfills the stated
condition be given. Then~$x = 0$ holds in the refinement~$B \defeq A/(x)$.
Hence~$1 = 0$ holds in~$B$. By elementary algebra, this means that~$x$ is
invertible in~$A$. Hence the conclusion holds for the singleton covering
of~$A$ by~$B_1 \defeq A$.

Within~$\Zar(\RR)$, we set~$\Delta \defeq \{ \varepsilon \? R \,|\,
\varepsilon^2 = 0 \}$. Then~$R$ has the following properties:
\begin{enumerate}
\item Law of cancellation: $\forall x \? R\_ \forall y \? R\_ \bigl((\forall
\varepsilon \? \Delta\_ x\varepsilon = y\varepsilon)
\Rightarrow x = y\bigr)$
\item Axiom of micro-affinity: $\forall f \? R^\Delta\_ \exists! a \? R\_
\forall \varepsilon \? \Delta\_ f(\varepsilon) = f(0) + a\varepsilon$
\end{enumerate}
The unique number~$a$ in the axiom of micro-affinity deserves to be
called~``$f'(0)$''; this is how we synthetically define the derivative in
synthetic differential geometry.

Having motivated the Zariski topos by the desire to devise a universe with
infinitesimals, the actual ontological status of the infinitesimal numbers in
the Zariski topos is more nuanced. The law of cancellation implies that,
within~$\Zar(\RR)$, it is not the case that zero is the only nilsquare number.
However, this doesn't mean that there actually \emph{is} a nilsquare number
in~$R$. In fact, any nilsquare number cannot be nonzero, as nonzero numbers are
invertible while nilsquare numbers are not. Hence any nilsquare number in~$R$
is \emph{not not} zero. This state of affairs is only possible in an
intuitionistic context.

\begin{rem}The ring~$R$ of smooth numbers does not coincide with the Cauchy reals,
the Dedekind reals or indeed any well-known construction of the reals
within~$\Zar(\RR)$. This observation explains why~$R$ can satisfy the law of
cancellation even though it is an intuitionistic theorem that the only
nilquadratic number in any flavor of the reals is zero.
\end{rem}

\subsection{Well-adapted models} The Zariski topos of~$\RR$ contains
infinitesimals in a satisfying manner. However it is not suited as a home for
synthetic differential geometry, a first indication being that in~$\Zar(\RR)$,
any function~$R \to R$ is polynomial. Hence important functions such as the
exponential function do not exist in~$\Zar(\RR)$. More comprehensively, the
Zariski topos is not a well-adapted model in the sense of the following
definition.

\begin{defn}\label{defn:well-adapted}A \emph{well-adapted model} of synthetic
differential geometry is a topos~$\E$ together with a ring~$R$ in~$\E$ such
that:
\begin{enumerate}
\item The ring~$R$ is a field.
\item The ring~$R$ validates the axiom of micro-affinity and several related
axioms.
\item There is a fully faithful functor~$i : \mathrm{Mnf} \to \E$ embedding the
category of smooth manifolds into~$\E$.
\item The ring~$R$ is coincides with~$i(\RR^1)$, the image of the real line
in~$\E$.
\end{enumerate}
\end{defn}

Is is then the culmination of a long line of research by several authors that
several well-adapted models of synthetic differential geometry exist. By the
conditions imposed by Definition~\ref{defn:well-adapted}, for any such
topos~$\E$ the following transfer principle holds: If~$f,g : \RR \to \RR$ are
smooth functions, then~$f' = g$ (in the ordinary sense of the derivative) if
and only if~$i(f)' = i(g)$ in~$\E$ (in the synthetic sense of the derivative).

Hence the nilquadratic infinitesimal numbers of synthetic differential geometry
may freely be employed as a convenient fiction when computing derivatives.
Because the theorem on the existence of well-adapted models has a constructive
proof, any proof making using of these infinitesimals may mechanically be
unwound to a (longer and more complex) proof which only refers to the ordinary
reals.

\subsection{On the importance of language}

XXX translation unwieldy, lens, view, ...

@@

* use the phrase "very rich and very deep" (or similarly) about set theory
* PA doesn't have models
* Sven(?)
* Set[T]
* free topos
* semantic view vs. syntactic view
* Kripke models
